---
title: "Indian Premier League T20 Analysis"
subtitle: "Data Science Professional Certificate Program from HarvardX - CYO"
author: "Riaz Mohamed Vellamparambil"
date: 02/21/2021
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(kableExtra)
```

\newpage
\tableofcontents


\newpage 

# Introduction
This is the submission for the choose-your-own project for the Harvard's multi-part Data Science Professional Certificate series. The goal of this project is to rank players and predict the winner of a match. T20 is a modern format of the original game of cricket where each ball and strike count. A team can acquire players through any of the three ways: the annual player auction, trading players with other teams during the trading windows, and signing replacements for unavailable players. Players sign up for the auction and also set their base price, and are bought by the franchise that bids the highest for them.


# About Cricket

https://en.wikipedia.org/wiki/Cricket

https://en.wikipedia.org/wiki/Indian_Premier_League

# Data Set

IPL ball by ball
https://www.kaggle.com/patrickb1912/ipl-complete-dataset-20082020?select=IPL+Ball-by-Ball+2008-2020.csv

IPL Matches
https://www.kaggle.com/patrickb1912/ipl-complete-dataset-20082020?select=IPL+Matches+2008-2020.csv


# Executive Summary:          

The goal of this project is to rank players by performance. This gives the IPL teams a pool of players as per caliber from which to pick from. 
We will also build a model to predict the winner of a match.


# Steps of Analysis

* **Data Preparation** - We will load the data, 
* **Data Exploration** - We will explore the data, plot histograms, create tables and graphs , As part of the data exploration process we will look at three key stats. Bowler Stats, Batsmen Stats and Team Stats
* **Modeling** -  We Will create two Models , one to create a pool of playes teams can choose from and predict match winner
* **Conclusion** - We will summarize the results of modeling


\newpage
# Data Preparation       


**Install and load libraries and MovieLens Dataset** 
To install the required packages, we use if(!require and load the packages from http://cran.us.r-project.org. Load Libraries and the IPL dataset


```{r message=FALSE, warning=FALSE}

# Install all needed libraries 

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(forcats)) install.packages("forcats", repos = "http://cran.us.r-project.org")
if(!require(ggplot2)) install.packages("ggplot2", repos = "http://cran.us.r-project.org")
if(!require(kableExtra)) install.packages("kableExtra" , repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(tidyr)) install.packages("tidyr", repos = "http://cran.us.r-project.org")
if(!require(stringr)) install.packages("stringr", repos = "http://cran.us.r-project.org")
if(!require(plotly)) install.packages("plotly", repos = "http://cran.us.r-project.org")
if(!require(ggthemes)) install.packages("ggthemes", repos = "http://cran.us.r-project.org")
if (!require(formattable)) install.packages("formattable", repos = "http://cran.us.r-project.org")
```

```{r message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)
library(caret)
library(kableExtra)
library(data.table)
library(dplyr)
library(tidyverse)
library(tidyr)
library(stringr)
library(forcats)
library(ggplot2)
library(plotly)
library(ggthemes)
library(formattable)
```

Import data sets from kaggle into R (I have downloaded the data from kaggle to my github repository)

```{r message=FALSE, warning=FALSE}
# read the .csv datafiles into R

ball_by_ball <- read.csv("https://raw.githubusercontent.com/riazvm/capstone-ipl/main/data/IPL-Ball-by-Ball-2008-2020.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)
matches <- read.csv("https://raw.githubusercontent.com/riazvm/capstone-ipl/main/data/IPL-Matches-2008-2020.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)

```


We will now inspect the data set to view column names 


Lets check how the data looks in the two data sets

```{r message=FALSE, warning=FALSE}
# inspect column nammes in the datasets (Can use colnames or names)
colnames(ball_by_ball)
names(matches)

# Lets check how the data looks in the two data sets
head(matches)
head(ball_by_ball)

```

The data matches data set shows all matches played between 2008 and 2020 between teams with a uique match id , we also see have other data that are described by each colum in the matches dataset
The ball_by_ball data set has each delivery bowled in the ipl matches and the match id corresponnds to the match id in the matches dataset. An over consists of 6 deliveries by a bowler


Let us check the data set for any anamolies , eg. duplicates, spellings etc. For this we will use the Factor function. Factor is a data structure used for fields that takes only predefined, finite number of values (categorical data). For example: a data field such as marital status may contain only values from single, married, separated, divorced, or widowed.


```{r message=FALSE, warning=FALSE}
# Let us check the data set for any anamolies , eg. duplicates, spellings etc. For this we will use the Factor 
levels(as.factor(ball_by_ball$batting_team))
levels(as.factor(ball_by_ball$bowling_team))
```

The following are the teams that play the IPL currently
Chennai Super Kings, Delhi Capitals, Gujarat Titans, Kolkata Knight Riders, Lucknow Super Giants, Mumbai Indians,
Punjab Kings, Rajasthan Royals, Royal Challengers Bangalore, Sunrisers Hyderabad. We will still need to keep the other teams that were playing the IPL previously to get the top players so the pool from which the teams can choose from is bigger.

Let is now conditionally replace the anomalies we see in the data

```{r message=FALSE, warning=FALSE}
#Let is now replace the anomalies we see in the correct data

# ball_by_ball data set

ball_by_ball$batting_team[ball_by_ball$batting_team == "Delhi Daredevils"] <- "Delhi Capitals"
ball_by_ball$batting_team[ball_by_ball$batting_team == "Kings XI Punjab"] <- "Punjab Kings"
ball_by_ball$batting_team[ball_by_ball$batting_team == "Gujarat Lions"] <- "Gujarat Titans"
ball_by_ball$batting_team[ball_by_ball$batting_team == "Gujarat Lions"] <- "Gujarat Titans"
ball_by_ball$batting_team[ball_by_ball$batting_team == "Rising Pune Supergiant"] <- "Lucknow Super Giants"
ball_by_ball$batting_team[ball_by_ball$batting_team == "Rising Pune Supergiants"] <- "Lucknow Super Giants"
ball_by_ball$batting_team[ball_by_ball$batting_team == "Pune Warriors"] <- "Lucknow Super Giants"

ball_by_ball$bowling_team[ball_by_ball$bowling_team == "Delhi Daredevils"] <- "Delhi Capitals"
ball_by_ball$bowling_team[ball_by_ball$bowling_team == "Kings XI Punjab"] <- "Punjab Kings"
ball_by_ball$bowling_team[ball_by_ball$bowling_team == "Gujarat Lions"] <- "Gujarat Titans"
ball_by_ball$bowling_team[ball_by_ball$bowling_team == "Gujarat Lions"] <- "Gujarat Titans"
ball_by_ball$bowling_team[ball_by_ball$bowling_team == "Rising Pune Supergiant"] <- "Lucknow Super Giants"
ball_by_ball$bowling_team[ball_by_ball$bowling_team == "Rising Pune Supergiants"] <- "Lucknow Super Giants"
ball_by_ball$bowling_team[ball_by_ball$bowling_team == "Pune Warriors"] <- "Lucknow Super Giants"

```


```{r message=FALSE, warning=FALSE}
# Check if we hav all the correct teams
levels(as.factor(ball_by_ball$batting_team))
levels(as.factor(ball_by_ball$bowling_team))

```

Let us check if the same type of errors exist in "matches" dataset also for team1, team2, winner, toss_winner and venue variables.


```{r message=FALSE, warning=FALSE}

# Lets look at the matches dataset for the same anomalies
levels(as.factor(matches$venue))

# The only Anamoly in this set is "M Chinnaswamy Stadium" & "M.Chinnaswamy Stadium"   

matches$venue[matches$venue == "M Chinnaswamy Stadium"] <- "M. A. Chidambaram Stadium"
matches$venue[matches$venue == "M. Chinnaswamy Stadium"] <- "M. A. Chidambaram Stadium"
matches$venue[matches$venue == "MA Chidambaram Stadium, Chepauk"] <- "M. A. Chidambaram Stadium"

levels(as.factor(matches$venue))
# Contiue other columns
levels(as.factor(matches$team1))


matches$team1[matches$team1 == "Delhi Daredevils"] <- "Delhi Capitals"
matches$team1[matches$team1 == "Kings XI Punjab"] <- "Punjab Kings"
matches$team1[matches$team1 == "Gujarat Lions"] <- "Gujarat Titans"
matches$team1[matches$team1 == "Gujarat Lions"] <- "Gujarat Titans"
matches$team1[matches$team1 == "Rising Pune Supergiant"] <- "Lucknow Super Giants"
matches$team1[matches$team1 == "Rising Pune Supergiants"] <- "Lucknow Super Giants"
matches$team1[matches$team1 == "Pune Warriors"] <- "Lucknow Super Giants"

levels(as.factor(matches$team2))

matches$team2[matches$team2 == "Delhi Daredevils"] <- "Delhi Capitals"
matches$team2[matches$team2 == "Kings XI Punjab"] <- "Punjab Kings"
matches$team2[matches$team2 == "Gujarat Lions"] <- "Gujarat Titans"
matches$team2[matches$team2 == "Gujarat Lions"] <- "Gujarat Titans"
matches$team2[matches$team2 == "Rising Pune Supergiant"] <- "Lucknow Super Giants"
matches$team2[matches$team2 == "Rising Pune Supergiants"] <- "Lucknow Super Giants"
matches$team2[matches$team2 == "Pune Warriors"] <- "Lucknow Super Giants"

levels(as.factor(matches$toss_winner))

matches$toss_winner[matches$toss_winner == "Delhi Daredevils"] <- "Delhi Capitals"
matches$toss_winner[matches$toss_winner == "Kings XI Punjab"] <- "Punjab Kings"
matches$toss_winner[matches$toss_winner == "Gujarat Lions"] <- "Gujarat Titans"
matches$toss_winner[matches$toss_winner == "Gujarat Lions"] <- "Gujarat Titans"
matches$toss_winner[matches$toss_winner == "Rising Pune Supergiant"] <- "Lucknow Super Giants"
matches$toss_winner[matches$toss_winner == "Rising Pune Supergiants"] <- "Lucknow Super Giants"
matches$toss_winner[matches$toss_winner == "Pune Warriors"] <- "Lucknow Super Giants"

levels(as.factor(matches$winner))

matches$winner[matches$winner == "Delhi Daredevils"] <- "Delhi Capitals"
matches$winner[matches$winner == "Kings XI Punjab"] <- "Punjab Kings"
matches$winner[matches$winner == "Gujarat Lions"] <- "Gujarat Titans"
matches$winner[matches$winner == "Gujarat Lions"] <- "Gujarat Titans"
matches$winner[matches$winner == "Rising Pune Supergiant"] <- "Lucknow Super Giants"
matches$winner[matches$winner == "Rising Pune Supergiants"] <- "Lucknow Super Giants"
matches$winner[matches$winner == "Pune Warriors"] <- "Lucknow Super Giants"

levels(as.factor(matches$winner))



```



\newpage
# Methods & Analysis - Data Exploration & Results:          
\vspace{12pt}

Let us start our exploration of data with finding out a few unique facts about our data.



```{r message=FALSE, warning=FALSE}
# Let us Check some unique sets

# Total players
n_distinct(ball_by_ball$player)

# Matches set 
matches %>% summarize(
  venue=n_distinct(matches$venue),
  teams=n_distinct(c(unique(matches$team1),unique(matches$team2))),
  matches_played=(tot_mat_played = n()),
  noresults=nrow(matches[is.na(matches$winner),])
)


# Total players, extras are the number of times that 
extras_data <- ball_by_ball %>% filter(ball_by_ball$extras_type %in% c("wides","noballs"))
nrow(extras_data)
```

**Conclusion:** We see a total of 507 players in the ball_by_ball data set. We also have 35 distinct venues, 
a total of 12 teams , 816 matches played and 4 ended up in no results which means it was a draw. And a total of 6616 extras in wide and noballs. The reason are looking at wides and no balls is because the bowler has to bowl 
an extra ball for every noball or wide bowled.


Now let us go into further analysis and check how many times a different run was scored by a player. Runs are
scored by batsmen and bowlers bowl deliveries.

```{r message=FALSE, warning=FALSE}
# Create a Player role

delivery <- ball_by_ball %>%
   gather(playerRole, player, batsman:bowler) %>%
  mutate(playerRole=as.factor(playerRole))

# Visual Representation: Let us look at how the runs are distributed as percentage

delivery %>% filter(playerRole == "batsman") %>%
  group_by(runs = total_runs) %>%
  summarize(count = n()) %>%
  mutate(distribution = percent(count / sum(count)))  %>% 
  ggplot(aes(x = runs , y = distribution*100))+
  ggtitle("Runs Percentage distribution")+
  xlab("Runs")+
  ylab("Distribution (%)")+
  geom_bar(stat = "identity",  fill = "orange", color = "black")+
  theme(plot.title = element_text(hjust = 1.0))

# Visual Representation: Let us see number of times a batsman had scored a different run 

delivery %>% 
  filter(playerRole == "batsman") %>%
  group_by(run = batsman_runs) %>%
  summarize(count = n()) %>%
  mutate(percent = percent(count / sum(count))) %>% 
  ggplot(aes(x = run , y = count))+
  ggtitle("Run Type Vs Count")+
  xlab("Runs")+
  ylab("Count")+
  geom_bar(stat = "identity",  fill = "orange", color = "black")+
  theme(plot.title = element_text(hjust = 1.0))



```

**Conclusion:** From the above visuals we do see that 7 runs is rare. So we will omit the 7 run. We also saw that there were were about 6616 extras.


Lets us now look if there is a correlation between a ball and the type of run scored. There are 20 overs in a T20
match for each innings. Each over has 6 deliveries with a total of 120 deliveries. If we take the average number of extras we will find that there is about 4 deliveries in an innings. 


```{r message=FALSE, warning=FALSE}
# Major run types scored on each ball and correlation between them
runs_and_batsman <- delivery %>%
  group_by(id, inning) %>%
  mutate(ball_no = 1:n()) %>% 
  ungroup() %>% 
  filter(playerRole == "batsman") %>%
  filter(batsman_runs != "" & batsman_runs != "7" & ball_no %in% 1: 124) %>%
  group_by(ball_no, batsman_runs) %>% 
  summarize(count=n())

runs_and_batsman

# Visualization on how runs stack up over the innings

runs_and_batsman %>%
  ggplot(aes(ball_no,count, col=factor(ball_no))) +
  geom_col() +
  facet_wrap( ~ batsman_runs) +
  scale_y_log10()+
  theme(axis.text.x = element_text(
    angle = 90,
    size = 10,
    hjust = 1
  ),
  legend.position = "none")

# Correlation
cor(runs_and_batsman$batsman_runs, as.numeric(runs_and_batsman$ball_no))

```

**Conclusion:** From the visualization we see that there is no correlation between the delivery and the runs scored .

We can see that the number of no runs scored decreases as the iniings progresses. We also see the 1s and 2s are 
steady throughout the innings.

Most teams try to bat out their innings by stabilizing thoughtout the innings


```{r message=FALSE, warning=FALSE}
# Run trends during the innings
delivery %>%
  group_by(id, inning) %>%
  mutate(ball_no = 1:n()) %>%
  ungroup() %>%
  filter (inning %in% 1:2 & playerRole == "batsman" & ball_no %in% 1:124) %>%
  group_by(ball_no=as.numeric(ball_no), runs = as.factor(batsman_runs)) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  arrange(ball_no) %>%
  top_n(5) %>%
  ggplot(aes(ball_no,count,col=runs, shape = runs)) +
  geom_point(size = 2)
```

**Conclusion:** From the visualization we see that 0s and 1's are the most prominent run types throughout the innings but are high in the first few overs and taper down towards the end of the innings. We also see the 4s and the 6's higher during both the initial part of the innings. This is because of the power play in T20 where
the field is restricted.


Lets look at how players performed during the IPL. Like all games the MVP dictates the player capability. We 
will look at the player_of_match field to pick the top 25 playes in the IPL so far.

\newpage
# Palyer Statistics : Batsman


```{r message=FALSE, warning=FALSE}
# top 25 players by player_of_match
matches %>%
  group_by(mvp=player_of_match) %>%
  summarize(player_of_match= n()) %>%
  arrange(desc(player_of_match)) %>%
  head(n = 25)
```

All teams look for batsman who are hard hitters , who can score boundaries at every ball. Let us see who 
are the players with maximum boundaries

```{r message=FALSE, warning=FALSE}
# Top 25 basmen with boundaries (boundary is a 4 or a 6)
delivery %>%
  filter(playerRole == "batsman" & batsman_runs == 6 | batsman_runs == 4) %>%
  group_by(player, batsman_runs) %>%
  summarize(n = n()) %>%
  spread(batsman_runs, n) %>%
  rename(sixes = `6`,fours = `4`) %>%
  select(sixes, fours) %>%
  arrange(desc(sixes)) %>%
  head(25)
```

Let us look at the top 25 batsmen with the maximum strike rate (Runs scored vs deliveries faced). The higher the better

```{r message=FALSE, warning=FALSE}
# Top 25 basmen 25 batsmen with the maximum strike rate (Runs scored vs deliveries faced)
delivery %>%
  filter(playerRole == "batsman") %>%
  group_by(player) %>%
  summarize(runs_scored=sum(batsman_runs), balls_faced=n(), strike_rate=sum(batsman_runs)/n()) %>%
  arrange(desc(strike_rate)) %>%
  head(25)
```



**Conclusion:** From the above analysis we know that the top 25 playes with the maximum runs are also the 
top boundary hitters. Strike rate cannot be taken to look into the top players because it is dependent on the number of balls faced. The top 25 in strike rate do not figure in the top players.

Now that we have seen the batsmen let us also look at the top bowlers in the IPL. A bowler can be concluded as 
a good bowler by the number of wickets taken and their econoomy rate (the less number of runs an over)

\newpage
# Palyer Statistics : Bowler

```{r message=FALSE, warning=FALSE}
# Let us check the the distict set of dimissal types
levels(as.factor(delivery$dismissal_kind))
  
```

Top bowlers with maximum number of wickets

```{r message=FALSE, warning=FALSE}
# Top 25 bowlers  with maximum number of wickets
delivery %>%
  filter(playerRole == "bowler" & 
      dismissal_kind != "" & dismissal_kind != "retired hurt" & dismissal_kind != "run out" & dismissal_kind != "run out"
  ) %>%
  group_by(player) %>%
  summarise(total_wickets=n())%>%
  arrange(desc(total_wickets)) %>%
  head(25)
  
``` 


Bowlers who conceded maximum runs

```{r message=FALSE, warning=FALSE}
# Top 25 bowlers  with maximum number of runs conceded
delivery %>%
  filter(playerRole == "bowler")  %>%
  group_by(player) %>%
  summarise(runsconceded=sum(batsman_runs), deliveries_bowled=n())%>%
  arrange(desc(runsconceded)) %>%
  head(25)
  
``` 


Lets check for the bowlers with good economy rate, which is the runs conceded vs the deliveries bowled The lower the economy rate the better

```{r message=FALSE, warning=FALSE}
# Top 25 bowlers  with best economy rate
delivery %>%
  filter(playerRole == "bowler") %>%
  group_by(player) %>%
  summarize(runs_conceded=sum(batsman_runs), deliveries_bowled=n(), economy_rate=sum(batsman_runs)/n()) %>%
  arrange(economy_rate) %>%
  head(25)
  
``` 

Lets check for the bowlers with good strike rate, which is the wickets taken vs the deliveries bowled The lower the strike rate the better


```{r message=FALSE, warning=FALSE}
# Top 25 bowlers  with best strike rate

t_wickets <- delivery %>%
  filter(playerRole == "bowler" & dismissal_kind != "" & dismissal_kind != "retired hurt" & dismissal_kind != "run out" & dismissal_kind != "run out") %>%
  group_by(player) %>%
  summarize(total_wickets = n())

top_max_deliveries <- delivery %>%
  filter(playerRole == "bowler") %>%
  group_by(player) %>%
  summarize(total_delivery = n()) %>%
  arrange(desc(total_delivery)) 
 

top_max_deliveries %>%
  full_join(t_wickets, by = "player") %>%
  mutate(strike_rate = total_delivery/ total_wickets) %>%
  arrange(strike_rate) %>%
  head(25)

  
``` 

**Conclusion:** From the above analysis there are bowlers  who show up in the top 25 because they have bowled lesser number of deliveries.  If the total number of wickets are taken into consideration as compared to the higest wicket takers its a few and these would not matter. We can eliminate the bowler strike rate during modeling our algorithm.


Let is now see how the top batsmen contributed to the overall IPL totals


```{r message=FALSE, warning=FALSE}
# The total runs in the IPL
total_ipl_runs <- delivery %>%
  filter(playerRole == "batsman") %>%
  summarize(total_runs_ipl = sum(total_runs)) 
total_ipl_runs

# The total runs by the top 25 batsman in the IPL
total_runs_by_top_25 <-  delivery %>%
  filter(playerRole == "batsman") %>%
  group_by(player) %>%
  summarize(tot_runs_top_batsmman = sum(batsman_runs))%>%
  arrange(desc(tot_runs_top_batsmman)) %>%
  head(n = 25)

sum(total_runs_by_top_25$tot_runs_top_batsmman)

# What is the percentage that they contributed

percent(sum(total_runs_by_top_25$tot_runs_top_batsmman)/total_ipl_runs$total_runs_ipl)

```

**Conclusion:** From the above analysis the top 25 batsmen contribute to 38% of the total runs in the IPL.


Let is now see how the top boowlers contributed to the overall IPL wickets


```{r message=FALSE, warning=FALSE}
# The total wickets in the IPL
total_ipl_wickets <- delivery %>%
  filter(playerRole == "bowler" & dismissal_kind != "") %>%
  summarize(overall_wickets = n())

total_ipl_wickets

# The wickets runs by the top 25 bowlers in the IPL
total_wickets_by_top_25 <- delivery %>%
  filter(playerRole == "bowler" & 
           dismissal_kind != "" & dismissal_kind != "retired hurt" & dismissal_kind != "run out" & dismissal_kind != "run out"
  ) %>%
  group_by(player) %>%
  summarise(total_wickets=n())%>%
  arrange(desc(total_wickets)) %>%
  head(25)

total_wickets_by_top_25


# What is the percentage that they contributed

percent(sum(total_wickets_by_top_25$total_wickets)/total_ipl_wickets)

```

**Conclusion:** From the above analysis the top 25 bowlers  contibute to 30% of the total wickets taken in the IPL

\newpage
# Team Statistics 

Let us analyze some Team statistics


```{r message=FALSE, warning=FALSE}
# Visual Total runs scored by each team

delivery %>% filter(playerRole == "batsman") %>%
  group_by(team = batting_team) %>%
  summarize(teamruns = sum(total_runs)) %>%
  ggplot(aes(x = teamruns , y = team))+
  ggtitle("Total Runs Per Team")+
  xlab("Runs")+
  ylab("Team")+
  geom_bar(stat = "identity",  fill = "red", color = "black")+
  theme(plot.title = element_text(hjust = 1.0))


# Match Winners

matches %>% 
  group_by(team = winner) %>%
  filter(winner != "") %>%
  summarize(count = n()) %>%
  ggplot(aes(x = count , y = team))+
  ggtitle("Total Wins Per Team")+
  xlab("Number of Wins")+
  ylab("Team")+
  geom_bar(stat = "identity",  fill = "blue", color = "black")+
  theme(plot.title = element_text(hjust = 1.0))

```

**Conclusion:** The top 3 teams are Mumbai Indians, Chennai Super Kinds and Kolkata Knight Riders , both in terms of wins and runs.

Does winning a toss matter to the team and does electing to field or bat matter to the team

```{r message=FALSE, warning=FALSE}
# Which team won how many tosses
toss_winners <- matches %>%
  group_by(toss_winner) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  rename(team = toss_winner, number_of_tosses_won = count)
toss_winners

# Matches played by teams
team1_played <- matches %>%
  group_by(team1) %>%
  summarize(count1 = n()) %>%
  arrange(team1) %>%
  rename(team = team1)

team2_played <- matches %>%
  group_by(team2) %>%
  summarize(count2 = n()) %>%
  arrange(team2) %>%
  rename(team = team2)

matches_team <- team1_played %>%
  full_join(team2_played, by = "team") 

matches_team
# Number of wins by eacxh team
winners <- matches %>%
  group_by(winner) %>%
  filter(winner != "") %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  rename(team = winner, matches_won = count)
winners

# Number of losses by each team

team1_lost <- matches %>%
  filter(winner != "") %>%
  filter(as.character(winner) != as.character(team1)) %>%
  group_by(team1) %>%
  summarize(count1 = n()) %>%
  arrange(team1) %>%
  rename(team = team1)
team2_lost <- matches %>%
  filter(winner != "") %>%
  filter(as.character(winner) != as.character(team2)) %>%
  group_by(team2) %>%
  summarize(count2 = n()) %>%
  arrange(team2) %>%
  rename(team = team2)
losers <- team1_lost %>%
  full_join(team2_lost, by = "team") %>%
  mutate(matches_lost = count1 + count2) %>%
  select(-count1,-count2) %>%
  arrange(desc(matches_lost))
losers

# Team matches , wins and losses
teams <- matches_team %>%
  right_join(toss_winners, by = "team")%>%
  right_join(winners, by = "team") %>%
  right_join(losers, by = "team") %>%
  mutate(matches_played = matches_won + matches_lost) 

teams
# Visualization of matches polayes vs won
teams %>%
  ggplot(aes(matches_played , matches_won, fill = team)) +
  geom_point(shape=21, size=5)

# Correlation between mamtches played to matches won
cor(teams$matches_played, teams$matches_won)
```

**Conclusion:** The more the matches the more wins. The correlation between matches played to matches won is also positive

How does matches won to tosses one correlate

```{r message=FALSE, warning=FALSE}
# Lets see the correlation between matches won and tosses won

teams %>%
  ggplot(aes(number_of_tosses_won, matches_won, fill = team)) +
  geom_point(shape=22, size = 5)
cor(teams$matches_won, teams$number_of_tosses_won)
```

**Conclusion:** Teams that won tosses mostly won matches as well. And there is a high correlation between the two

Let us now check the decision to bat or field after winning the toss affects the wins


```{r message=FALSE, warning=FALSE}

# Toss wins and toss decisions

toss_winners <- matches %>%
  group_by(toss_winner) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  rename(team = toss_winner, number_of_tosses_won = count)


toss_wins_results <- matches %>%
  filter(winner != "") %>%
  mutate(match_result = ifelse(toss_winner==winner, "win", "loss")) %>%
  group_by(toss_winner, toss_decision, match_result) %>%
  summarize(match_result_count = n()) 

toss_wins_results

# Visualization : Effect of toss wins vs decision to bat or field
toss_wins_results %>% 
  ggplot(aes(toss_winner, match_result_count, col = toss_decision, size = match_result)) +
  geom_point()  +
  ggtitle("Toss Decision Vs Winner")+
  xlab("Toss_Winner")+
  ylab("Matches")+
  theme(axis.text.x = element_text(
    angle = 50,
    size = 6,
    hjust = 1
  ),
  legend.position = "right")
```

**Conclusion:** From the plot we see that the teams that have elected to field first after winning the toss have
won more games  (bigger circles). The decision to win the toss thus effects the result of a match.


Let us check if venue plays a part in the result of a game


At this point, let us check how many runs were scored batting first and how many were scored batting second in totals and run type wise.

```{r message=FALSE, warning=FALSE}

# Venue effect on team wins
matches %>%
  filter(winner != "") %>%
  group_by(venue, winner) %>%
  summarize(count = n()) %>%
  ggplot(aes(venue, count, group = winner)) +
  geom_point() +
  ggtitle("Venu Vs Wins")+
  xlab("Venue")+
  ylab("Win Count")+
  geom_line(aes(col = winner), size = 1) +
  theme(axis.text.x = element_text(
    angle = 50,
    size = 6,
    hjust = 1
  ),
  legend.position = "right", legend.title.align=0)
```

**Conclusion:** From the plot we can see that the home venues result in more wins ffor the teams as inidicated by the peaks

We have enough data to Start our model  


\newpage
# Model 1 : Build a model to rank the best players depending on his value

Factors determining the value of a player
    1)  High Strike Rate (Runs/deliveries faced) for a batsman
    2)  Low Economy Rate (Runs given/ deliveries bowled)
    3)  Socre against top bowlers and economy against top batsmen
    4)  Runs Scored by the team and wickets dismissed


\newpage
### High Strike Rate (Runs/deliveries faced) for a batsman
### Low Economy Rate (Runs given/ deliveries bowled)

\vspace{12pt}

Let us determine the batsmen with highest strike rates. For this let is create a table that contains the averages


```{r message=FALSE, warning=FALSE}

# batsmen stats
batsmen_avgs <- delivery %>%
  filter(playerRole == "batsman") %>%
  group_by(player) %>%
  summarize(total_deliveries = n(), tot_runs = sum(batsman_runs)) %>%
  summarize (
    avg_deliveries = mean(total_deliveries),
    avg_runs = mean(tot_runs),
    median_deliveries = median(total_deliveries),
    median_runs = median(tot_runs)
  )
as_tibble(batsmen_avgs)




# bowler stats
bowler_avgs <- delivery %>%
  filter(playerRole == "bowler") %>%
  group_by(player) %>%
  summarize(total_deliveries = n(), tot_runs = sum(batsman_runs)) %>%
  summarize (
    avg_deliveries = mean(total_deliveries),
    avg_runs = mean(tot_runs),
    median_deliveries = median(total_deliveries),
    median_runs = median(tot_runs)
  )
as_tibble(bowler_avgs)

```


Model: Reguliration - We will calculate the strike rate and economy rates using regularization. As we have already seen during our analysis the strike rate can be high but most of the players with high strike rates do not figure into the top 25 players. If they face just 2 deliveries and hit a boundary in the first their strike rate can be very high. Hence we adjust this by using median as the Error. 


```{r message=FALSE, warning=FALSE}


strike_rates_regularized <- delivery %>%
  filter(playerRole == "batsman") %>%
  group_by(player) %>%
  summarize(batsman_reg_str_rate = (sum(batsman_runs) + batsmen_avgs$median_runs) / (n() +
                                                                batsmen_avgs$median_deliveries)) %>%
  arrange(desc(batsman_reg_str_rate)) 

strike_rates_regularized %>%
  head(25)


economy_rates_regularized <- delivery %>%
  filter(playerRole == "bowler") %>%
  group_by(player) %>%
  summarize(bowler_reg_eco_rate = (sum(batsman_runs) + bowler_avgs$median_runs) / 
              (n() + bowler_avgs$median_deliveries)) %>%
  arrange(bowler_reg_eco_rate)
economy_rates_regularized %>%
  head(25)

```

 Now let us merge the two sets and get the top players and rank them. When we merge these two sets we will have a few not applicable values , the reason is because a batsman may not be a bowler and vice versa. If a Batsman is nt a bowler we will use the avg balls and for bwlers who are not batsman we will use the median runs

```{r message=FALSE, warning=FALSE}
# Top rate players based on strike rates & economy rates
top_players <- strike_rates_regularized %>%
  full_join(economy_rates_regularized, by = "player") %>%
  mutate(batsman_reg_str_rate = replace_na(
    batsman_reg_str_rate,
    batsmen_avgs$median_runs / batsmen_avgs$avg_deliveries
  )) %>%
  mutate(bowler_reg_eco_rate = replace_na(bowler_reg_eco_rate, bowler_avgs$avg_runs /
                                     bowler_avgs$median_deliveries)) %>%
  mutate(value_of_player = 100 * (batsman_reg_str_rate + 1 / bowler_reg_eco_rate)) 
top_players %>%
  arrange(desc(value_of_player)) %>%
  select(player, value_of_player) %>%
  mutate(rank = row_number()) %>%
  head(100) %>% 
  knitr::kable()
```

This table gives us the top players  bowlers , batsmen and allrounders and theur value.

We will now introduce an anamoly which is how the top bowlers and batsmen performed against each other

```{r message=FALSE, warning=FALSE}

# Top 25 batsman and bowlers
top_batsmen <- strike_rates_regularized %>%
  head(25)
top_batsmen
top_bowlers <- economy_rates_regularized %>%
  head(25)
top_bowlers


# batsmen strike rate against top_bowlers bowlers
# we will use the original set

top_25_bat_strikerate<- ball_by_ball %>%
  filter(bowler %in% top_bowlers$player) %>%
  group_by(player = batsman) %>%
  summarize(strike_rate = (sum(batsman_runs) + batsmen_avgs$median_runs) / (n() +batsmen_avgs$avg_deliveries)) %>%
  arrange(desc(strike_rate))
top_25_bat_strikerate %>%
  mutate(rank = row_number())%>%
  head(25) 


# bowler ecorate rate against top_bowlers bowlers
# we will use the original set

top_25_bow_ecorate <- ball_by_ball %>%
  filter(batsman %in% top_batsmen$player) %>%
  group_by(player = bowler) %>%
  summarize(eco_rate = (sum(batsman_runs) + bowler_avgs$avg_runs) / (n() +bowler_avgs$median_deliveries)) %>%
  arrange(eco_rate)
top_25_bow_ecorate %>%
  mutate(rank = row_number())%>%
  head(25) 
```

 Now let us merge the two sets and get the top players and rank them. When we merge these two sets we will have a few not applicable values , the reason is because a batsman may not be a bowler and vice versa. If a Batsman is not a bowler we will use the avg balls and for bowlers who are not batsman we will use the median runs


```{r message=FALSE, warning=FALSE}
# Top class players based on strike rates & economy rates against each other
top_class_players <- top_25_bat_strikerate %>%
  full_join(top_25_bow_ecorate, by = "player") %>%
  mutate(strike_rate = replace_na(strike_rate,
    batsmen_avgs$median_runs / batsmen_avgs$avg_deliveries
  )) %>%
  mutate(eco_rate = replace_na(eco_rate, bowler_avgs$avg_runs /
                                            bowler_avgs$median_deliveries)) %>%
  mutate(value_of_player = 100 * (strike_rate + 1 / eco_rate)) 


top_class_players %>%
  arrange(desc(value_of_player)) %>%
  select(player, value_of_player, strike_rate,eco_rate) %>%
  mutate(rank = row_number()) %>%
  head(100) %>% 
  knitr::kable()
```



```{r message=FALSE, warning=FALSE}
# Top class players based on strike rates & economy rates against each other
top_class_players <- top_25_bat_strikerate %>%
  full_join(top_25_bow_ecorate, by = "player") %>%
  mutate(strike_rate = replace_na(strike_rate,
    batsmen_avgs$median_runs / batsmen_avgs$avg_deliveries
  )) %>%
  mutate(eco_rate = replace_na(eco_rate, bowler_avgs$avg_runs /
                                            bowler_avgs$median_deliveries)) %>%
  mutate(value_of_player = 100 * (strike_rate + 1 / eco_rate)) 


top_class_players %>%
  arrange(desc(value_of_player)) %>%
  select(player, value_of_player, strike_rate,eco_rate) %>%
  mutate(rank = row_number()) %>%
  head(100) %>% 
  knitr::kable()
```




We will now create the pool of 200 top players based on the above two models. We are taking into account players ranked by strikerate and economy and also how they performed against each other. There maybe other anomolies we can introduce to this to arrive at the best pool of players but we will restrict our findings to these two.



```{r message=FALSE, warning=FALSE}
# Top pool players based on strike rate and economy for bowlers and their performance with each other
# we will again join the two sets and replace the strike rate and economy rate as previously done
world_top_players <- top_players %>%
  select(-value_of_player) %>%
  full_join(top_class_players, by = "player") %>%
  mutate(strike_rate = replace_na(strike_rate,
                                  batsmen_avgs$median_runs / batsmen_avgs$avg_deliveries
  )) %>%
  mutate(eco_rate = replace_na(eco_rate, bowler_avgs$avg_runs /
                                 bowler_avgs$median_deliveries)) %>%
  mutate(player_value = 100 * ((batsman_reg_str_rate + strike_rate) + 1 /
                                 (bowler_reg_eco_rate + eco_rate)))  %>%
  select(player, value_of_player) %>%
  arrange(desc(value_of_player)) %>%
  mutate(rank = row_number())
world_top_players %>%
  head(200)
```


\newpage
# Model 2: Predict the winner of a Match

From our analysis of the matches we have seen the following factors matter

    1)  venue, toss, toss decision, team battiing first and the team batting next
    2)  the data set we have is extremely small

To predict a match winner it is known that the (KNN and Random Forest), with 49.77% accuracy (logistic Regression), almost 17% better than a random decision (benchmark) which has 33%. We will also look at the Naive Bayes methodology.

Naive Bayes -- https://en.wikipedia.org/wiki/Naive_Bayes_classifier
Logistic Regression -- https://towardsdatascience.com/introduction-to-logistic-regression-66248243c148#:~:text=Logistic%20regression%20is%20a%20classification,Fraud%2C%20Tumor%20Malignant%20or%20Benign.
KNN Algorithm (Nearest Neighbour) -- https://en.wikipedia.org/wiki/K-nearest_neighbors_algorithm
Random Forest  -- https://en.wikipedia.org/wiki/Random_forest






\newpage
# Data Sets: train and test datasets      

Create the training and test sets

```{r message=FALSE, warning=FALSE}
# Create a new dataset. We are going to be taking matches as our base data set to work out of
# Let us remove  Gujarat Titans and Kochi Tuskers Kerala because they  no more take part in the IPL
best_teams <- teams %>% arrange(desc(matches_played)) %>%
  head(8)
best_teams
match_predictor_ds <- matches %>%
  select(team1,team2, winner, venue, toss_winner, toss_decision) %>%
  filter(winner != ""  & team1 %in% best_teams$team & team2 %in% best_teams$team)
any(is.na(match_predictor_ds))
summary(match_predictor_ds)
match_predictor_ds

n_distinct(match_predictor_ds$team1)

levels(as.factor(match_predictor_ds$team1))

match_predictor_ds$team1[match_predictor_ds$team1 == "Mumbai Indians"] <- "Mumbai"
match_predictor_ds$team1[match_predictor_ds$team1 == "Delhi Capitals"] <- "DC"
match_predictor_ds$team1[match_predictor_ds$team1 == "Kolkata Knight Riders"] <- "KKR"
match_predictor_ds$team1[match_predictor_ds$team1 == "Royal Challengers Bangalore"] <- "RCB"
match_predictor_ds$team1[match_predictor_ds$team1 == "Punjab Kings"] <- "Punjab"
match_predictor_ds$team1[match_predictor_ds$team1 == "Chennai Super Kings"] <- "CSK"
match_predictor_ds$team1[match_predictor_ds$team1 == "Sunrisers Hyderabad"] <- "Hyd"
match_predictor_ds$team1[match_predictor_ds$team1 == "Rajasthan Royals"] <- "Rajasthan"


levels(as.factor(match_predictor_ds$team2))
match_predictor_ds$team2[match_predictor_ds$team2 == "Mumbai Indians"] <- "Mumbai"
match_predictor_ds$team2[match_predictor_ds$team2 == "Delhi Capitals"] <- "DC"
match_predictor_ds$team2[match_predictor_ds$team2 == "Kolkata Knight Riders"] <- "KKR"
match_predictor_ds$team2[match_predictor_ds$team2 == "Royal Challengers Bangalore"] <- "RCB"
match_predictor_ds$team2[match_predictor_ds$team2 == "Punjab Kings"] <- "Punjab"
match_predictor_ds$team2[match_predictor_ds$team2 == "Chennai Super Kings"] <- "CSK"
match_predictor_ds$team2[match_predictor_ds$team2 == "Sunrisers Hyderabad"] <- "Hyd"
match_predictor_ds$team2[match_predictor_ds$team2 == "Rajasthan Royals"] <- "Rajasthan"

levels(as.factor(match_predictor_ds$winner))

match_predictor_ds$winner[match_predictor_ds$winner == "Mumbai Indians"] <- "Mumbai"
match_predictor_ds$winner[match_predictor_ds$winner == "Delhi Capitals"] <- "DC"
match_predictor_ds$winner[match_predictor_ds$winner == "Kolkata Knight Riders"] <- "KKR"
match_predictor_ds$winner[match_predictor_ds$winner == "Royal Challengers Bangalore"] <- "RCB"
match_predictor_ds$winner[match_predictor_ds$winner == "Punjab Kings"] <- "Punjab"
match_predictor_ds$winner[match_predictor_ds$winner == "Chennai Super Kings"] <- "CSK"
match_predictor_ds$winner[match_predictor_ds$winner == "Sunrisers Hyderabad"] <- "Hyd"
match_predictor_ds$winner[match_predictor_ds$winner == "Rajasthan Royals"] <- "Rajasthan"


levels(as.factor(match_predictor_ds$toss_winner))
match_predictor_ds$toss_winner[match_predictor_ds$toss_winner == "Mumbai Indians"] <- "Mumbai"
match_predictor_ds$toss_winner[match_predictor_ds$toss_winner == "Delhi Capitals"] <- "DC"
match_predictor_ds$toss_winner[match_predictor_ds$toss_winner == "Kolkata Knight Riders"] <- "KKR"
match_predictor_ds$toss_winner[match_predictor_ds$toss_winner == "Royal Challengers Bangalore"] <- "RCB"
match_predictor_ds$toss_winner[match_predictor_ds$toss_winner == "Punjab Kings"] <- "Punjab"
match_predictor_ds$toss_winner[match_predictor_ds$toss_winner == "Chennai Super Kings"] <- "CSK"
match_predictor_ds$toss_winner[match_predictor_ds$toss_winner == "Sunrisers Hyderabad"] <- "Hyd"
match_predictor_ds$toss_winner[match_predictor_ds$toss_winner == "Rajasthan Royals"] <- "Rajasthan"

match_predictor_ds$team1
match_predictor_ds$team2
# Create a Train Set and a Test Set

set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
options(digits=3)

test_index <- createDataPartition(y = match_predictor_ds$winner, times = 1, p = 0.1, list = FALSE)

# create a test & train set
train <- match_predictor_ds[-test_index,]
temp <- match_predictor_ds[test_index,]

# Matching team1,team2,venue,toss_decision and toss_winner in both train and test sets
test <- temp %>%
  semi_join(train, by = "team1") %>%
  semi_join(train, by = "team2") %>%
  semi_join(train, by = "venue") %>%
  semi_join(train, by = "toss_decision") %>%
  semi_join(train, by = "toss_winner") 
# Add rows removed from temp set back into train set
removed <- anti_join(temp, test)
train <- rbind(train, removed) 
rm(removed, temp, test_index)

#Let us explore the train data set

train %>% as_tibble()

test %>% as_tibble()
```


\newpage
# Algorithms and Fit

F score : In statistical analysis of binary classification, the F-score or F-measure is a measure of a test's 
accuracy. It is calculated from the precision and recall of the test, where the precision is the number of 
true positive results divided by the number of all positive results, including those not identified correctly, 
and the recall is the number of true positive results divided by the number of all samples that should have 
been identified as positive. Precision is also known as positive predictive value, and recall is also known 
as sensitivity in diagnostic binary classification.

We are not worried abut negative outcomes and hence going with an F score.
F1 score: https://towardsdatascience.com/the-f1-score-bec2bbc38aa6 , we will use F1 score.
Naive Bayes

```{r message=FALSE, warning=FALSE}
# Since all the variables in the  dataset are variables that deermine he winner we will calulcae the F1 score 
# for all classes
naive <- train(winner ~ ., method = "naive_bayes", data = train)
predicion_test_naive <- predict(naive, test)
f1_score_nb <- confusionMatrix(predicion_test_naive, as.factor(test$winner))$byClass[,"F1"]
f1_score_nb <- as.data.frame(t(f1_score_nb)) %>% mutate(averageF1Score = rowMeans(.))
# Table
# Make column names more readable
colnames(f1_score_nb) = gsub("Class: ", "", colnames(f1_score_nb))
# F1 table for different models
f1_model_table <- data.frame(Model = "Naive") %>% bind_cols(f1_score_nb)
f1_model_table %>% knitr::kable()
```
We can see from the observation that the prediction oes nt satisfy all teams and hence the Naive Bayes algorithm cannot be used in this case


Logistic Regression 


```{r message=FALSE, warning=FALSE}
# Logistic Regression

logistic_regression <- train(winner ~ ., method = "lda", data = train)
predicion_test_lr <- predict(logistic_regression, test)
f1_score_lr <- confusionMatrix(predicion_test_lr, as.factor(test$winner))$byClass[,"F1"]
f1_score_lr <- as.data.frame(t(f1_score_lr)) %>% mutate(averageF1Score = rowMeans(.))
f1_score_lr

# Table
colnames(f1_score_lr) = gsub("Class: ", "", colnames(f1_score_lr))

f1_model_table <- bind_rows(f1_model_table,
                      data.frame(Model = "LDA") %>% bind_cols(f1_score_lr))
f1_model_table %>% knitr::kable()
```


Random Forest


```{r message=FALSE, warning=FALSE}
# Random Forest (multiple trees)

trainctrl <- trainControl(method="cv")
random_forest <- train(winner ~ ., method = "rf", data = train,  trControl=trainctrl)
predicion_test_rf <- predict(random_forest, test)
f1_score_rf <- confusionMatrix(predicion_test_rf,  as.factor(test$winner))$byClass[,"F1"]
f1_score_rf <- as.data.frame(t(f1_score_rf)) %>% mutate(averageF1Score = rowMeans(.))
f1_score_rf

colnames(f1_score_rf) = gsub("Class: ", "", colnames(f1_score_rf))
f1_model_table <- bind_rows(f1_model_table,
                            data.frame(Model = "RF") %>% bind_cols(f1_score_rf))
f1_model_table %>% knitr::kable()

```

Nearest Neighbour


```{r message=FALSE, warning=FALSE}
# KNN (Nearest neighbour)

knn <- train(winner ~ ., method = "knn", data = train,)
predicion_test_knn <- predict(knn, test)
f1_score_knn <- confusionMatrix(predicion_test_knn, as.factor(test$winner))$byClass[,"F1"]
f1_score_knn <- as.data.frame(t(f1_score_knn)) %>% mutate(averageF1Score = rowMeans(.))
f1_score_knn

# Table
colnames(f1_score_knn) = gsub("Class: ", "", colnames(f1_score_knn))
f1_model_table <- bind_rows(f1_model_table,
                      data.frame(Model = "KNN") %>% bind_cols(f1_score_knn))
f1_model_table %>% knitr::kable()
```


\newpage
# Conclusion: 

We have created a pool of players considering different factors from which IPL teams can pick from. There are always biases that can be added to the analysis Eg. What if a top player moved teams , that may change the winner of a game. 
We also see from the analysis that LDA gives us the best results in terms of predicting a winner. Random forest as it considers multiple trees is a little slow compared to the other algorithms. KNN and Random forest results are nearly similar.

I also want to take this oppertunity to thanks all my peers , staff and the excellent teachers at the program.




